Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-22 manuscript.rmd
added Brent's edits. other edits

- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data. Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{x,~z}$ by depth $z$in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$) linking data on the same for 21 species with community-scale softer traits. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$). 
- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate. Deeper ERD species spent less time below critical hydraulic thredholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome based on observed ranges of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in testing this mechanism community-wide has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic concentration in plant tissue to it's gradient vertically belowground. Such data is available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].

\
Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$). A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.

@@

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis. Specifically, in this forest (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not usre if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

@@

## The Study Site
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560 hectare Barro Colorado Island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April  (≤ 100 mm rain/month). The entire Island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit:1998395; @Hubbell:1999; @Anderson-Teixeira:2014].

@@

\
Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note). ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in the unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as

@@

\
We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter waws chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES. This figure will be reformatted to compose individual figures into one."}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

@@

For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7). We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil. We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present. Fig. S\@ref(fig:iso-data). We also removed *Guapira* *spp*, whose xylem sap value fell in the uncertain ranges of soil depth.

\
Model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (dropped one species with average abundance of 9 trees). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~{1}~ and cen~{2}~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~{1}~ and dead in cen~{2}~, respectively, and t is the duration based on mean dates of cen~{1}~ and cen~{2}~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{i}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~Leaf}$.

@@

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island. See Table S2 for scientific names of species codes")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. Each circle resepresents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species. In this moist forest ($\Psi_{z}$) for depths at and below 1.7 m remained above ($\Psi_{crit}$) of the most sensitive tree species through all the dryseason and drought periods over 1990-2015, period over which ($\Psi_{z}$) was modelled. We thus show, for the first time, that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axism, by demonstrating a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tropical forest community.

@@

- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients from soil through plant tissues (roots-stems-leaves) to atmosphere continuum (SPAC) [@Johnson:2017]. However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves. (Note that In this forest LMA and WSG did not scale with stem vulnerability curves). Water storage (capacitance) should also be considered.

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme droughts by census interval and some representative depths"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```

>>>>

- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data. Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{x,~z}$ by depth $z$in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$) linking data on the same for 21 species with community-scale softer traits. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$). 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora:2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].

\
Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be a challenge. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2017]. A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water[@Nardini:2015]. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated becuase of lack of data on rooting/water-sourcing depths, or related hydraulic traits.

@@

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis. Specific hypotehses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

@@

## The Study Site
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit:1998395; @Hubbell:1999; @Anderson-Teixeira:2014].

@@

\
Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note). ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as

@@

\
We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES. This figure will be reformatted to compose individual figures into one."}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

@@

For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7). We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for the dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil. We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present. Fig. S\@ref(fig:iso-data). We also removed *Guapira* *spp*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.

\
The model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]"). See Table S2 for scientific names corresponding to species codes.}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~Leaf}$.

@@

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island. See Table S2 for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on an acquisitive-conservative strategy axis by demonstrating a link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes. On the aquisitive end were species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable, abundant source of water at depth, which actualised to a lower associated mortality-risk, whilst at the conservative end were species with inefficient, but safer xylem, associated with a water environment of high variability.

Acquisitive strategy is thought to be optimised for high resource environments with high performance outcome and a large decline in performance under stresses or low resource environment; whilst conservative strategy have a moderate performance in high resource environment that declines with decreasing resources, while retaining a competitive advantage over acquisitives under lower resources.

In other contexts aquisitive-conservative trait trade-off is associated with a growth-survival trade-off, with aquisitive species achieving higher potential growth at the cost of lower survival, but in case of ERD axis, aquisitiveness did not relate Normally Our results highlight the need.. to higher growth, but

## Trait-Trait Coordination

-- trade-offs among differential investment in hydraulic efficiency and safety
here links with trade-off in water-stress vertically belowground
-- unexplored (hydraulic capacitance, stomatal control, photosynthetic capacity, high temperature tolerance)

## Trait & Demography

## Environment, Trait & Demography

Mortality could be immediate or may result after decades of drought occurrence.
Mortality could occur over

@@

- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients over the soil - plant (roots, stems, leaves) - atmosphere continuum (SPAC) [@Johnson:2017; @Christoffersen 2017]. However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development would be needed to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves. (Note that In this forest LMA and WSG did not scale with stem vulnerability curves). Water storage (capacitance) should also be considered.

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths (0.5, 1 and 1.7 m depicted in different colors). Mean soil water potential ($Psi_z$}) for a given DOY and depth z are shown as lines. Filled areas show extreme drought periods, when $Psi_z$ declined below 2.5 percentile of $Psi_z$ for the given DOY and depth z in an individual year within the cenus intervals. Eextreme drought periods from multiple years may be overlaid within a panel."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```
